
\section{Introduction}

Let $\mathbf{F}\in\mathbb{K}\left[x\right]^{m\times n}$ be a matrix
of polynomials over a field $\mathbb{K}$ with $m\le n$. A (right)
nullspace basis $\mathbf{N}$ of $\mathbf{F}$ is said to be minimal
if it has the minimal possible column degrees among all right nullspace
bases. More generally, for a shift $\vec{s}=[s_{1},\dots,s_{n}]\in\mathbb{Z}^{n}$,
a nullspace basis $\mathbf{N}$ is said to be $\vec{s}$-minimal if
it has the minimal possible $\vec{s}$-column degrees among all nullspace
bases, or equivalently, the column degrees of 
\[
x^{\vec{s}}\mathbf{N}=\begin{bmatrix}s_{1}\\
 & \ddots\\
 &  & s_{n}
\end{bmatrix}\mathbf{N}
\]
 are the minimal possible among all nullspace bases of $\mathbf{F}$.

In this paper we present an algorithm for computing a minimal nullspace
basis with a cost of $O^{\sim}\left(n^{\omega}d\right)$. A key ingredient
of this algorithm is another algorithm, also presented in this paper,
for computing a $\vec{s}$-minimal shifted nullspace basis of an input
matrix with row dimension $m$ and column dimension $n$ satisfy $n\in\Theta(m)$,
where the entries of the shift $\vec{s}$ bound the column degrees
of $\mathbf{F}$ component-wise, that is, each entry $s_{i}$ is greater
than or equal to the degree of the $i$-th column of $\mathbf{F}$.
For example, the shift $\vec{s}$ can be simply the column degrees
of $\mathbf{F}$. If $\xi$ is the sum of the entries of $\vec{s}$,
a $\vec{s}$-minimal nullspace basis can be computed with a cost of
$O^{\sim}(n^{\omega-1}\xi)$ field operations.

A key component of the algorithm is order basis (also known as minimal
approximant basis) computation, which are done using the algorithms
from \citet{Giorgi2003} and \citet{za2009}. We use order basis computation
to compute a partial nullspace basis, which also reduces the column
dimension of the problem. The problem can then be separated to two
subproblems of smaller row dimensions, which can then be handled in
the same way as the original problem. 

The remaining paper is structured as follows. Basic definitions and
properties of nullspace bases and order bases are given in the next
section. \prettyref{sec:transform} provides an extension to Storjohann's
transformation to allow higher degree basis elements to be computed.
Based on this new transformation, \prettyref{sec:Order-Basis-Computation}
establishes a link between two Storjohann transformed problems of
different degrees, from which an recursive method and then an iterative
algorithm are derived. The time complexity is analyzed in the next
section. After this, \prettyref{sec:Unbalanced-Shift} describes an
algorithm which handles problems with a type of unbalanced shift.
This is followed by a conclusion along with a description for topics
for future research. 
